---
title: "Lab 03 - Data Wrangling"
author: Ives He
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## 0 load the package 
```{r}
library(tidyverse)
```

## 1 create a tibble 

```{r}
# Load the data
firm_data <- read_csv("https://raw.githubusercontent.com/PYHkkk/FIN-550/main/firm.csv")

# First six rows
head(firm_data)

# Number of rows
nrow(firm_data)

# Number of columns
ncol(firm_data)

# Column names
colnames(firm_data)
```

## 2 subset data 

### 2.1 one column 
```{r}
# Selecting one column, e.g., 'TIC'
firm_data %>% select(TIC)
```

### 2.2 three columns 
```{r}
# Selecting three columns, e.g., 'TIC', 'FYEAR', and 'NI'
firm_data %>% select(TIC, FYEAR, NI)
```

### 2.3 four rows and three columns
```{r}
# Selecting first four rows and three columns
firm_data %>% select(TIC, FYEAR, NI) %>% slice(1:4)
```

## 3 filter rows  

### 3.1 observations that belong to AAPL in 2015 
```{r}
firm_data %>% filter(TIC == "AAPL" & FYEAR == 2015)
```

### 3.2 observations that have the minimum NI value 
```{r}
firm_data %>% filter(NI == min(NI, na.rm = TRUE))
```

### 3.3 select columns of observations that belong to AAPL in 2015   
```{r}
firm_data %>% filter(TIC == "AAPL" & FYEAR == 2015) %>% select(TIC, FYEAR, NI)
```

## 4 arrange rows 

### 4.1 ascending order of NI
```{r}
firm_data %>% arrange(NI)
```

### 4.2 descending order of NI
```{r}
firm_data %>% arrange(desc(NI))
```

### 4.3 ascending order of TIC and FYEAR, and descending order of NI
```{r}
firm_data %>% arrange(TIC, FYEAR, desc(NI))
```

## 5 create new variables  

### 5.1 return on assets 
```{r}
firm_data <- firm_data %>% mutate(ROA = NI / AT)
firm_data
```

### 5.2 net income bin numbers  
```{r}
# step1: identify thresholds of 4 bins with an equal number of observations 
ni_bins <- quantile(firm_data$NI, probs = seq(0, 1, length.out = 5), na.rm = TRUE)
# step2: remove the names of a named vector  
ni_bins <- unname(ni_bins)
# step3: bin NI  
firm_data <- firm_data %>% mutate(NI_bin = cut(NI, breaks = ni_bins, include.lowest = TRUE))
firm_data
```

### 5.3 log of assets 
```{r}
firm_data <- firm_data %>% mutate(Log_AT = log(AT))
firm_data
```

### 5.4 drop a column 
```{r}
firm_data <- firm_data %>% select(-Log_AT)
firm_data
```

