---
title: "Lecture 2: Introduction to R programming"
output:
  html_document:
---



# Try it: list/vector subsetting example

- Open the R markdown file for this lecture (`lecture-02-tryit.Rmd`)

- Execute the code below to create a vector and a list

```{r}
# Define an atomic vector
x_vector = c(2, 4, 6, 8)

# Define a named list
x_list = list(Name = c("Riley", "League"), 
              Subject = "FIN", Number = 550)
x_list
```



- Use single brackets `[` to select one or more elements of a vector or list

```{r, eval=FALSE}
# Positive integers select elements in given positions
x_vector[1]
x_vector[c(1,4)]
# Negative integers drop elements in given positions
x_vector[-1]
x_vector[c(-1,-4)]
# Logical vectors select elements with a corresponding value of TRUE
x_vector[c(TRUE, FALSE, FALSE, FALSE)]
# When subsetting a list, `[` returns a list containing the selected elements
x_list[1]

```

- Use double brackets `[[` or the dollar sign `$` to extract a single element

```{r, eval=FALSE}
# Positive integer extracts element in given position
x_vector[[1]]
# For named lists, can extract an element by name
x_list[[1]]
x_list[["Name"]]
# Using `[[` on a list returns the selected element itself
# Using `[` on a list returns a list with the selected element
# Note: you can confirm this by using `typeof()`
typeof(x_list[1])
typeof(x_list[[1]])
```


# Try it: conditional statements (if, if/else)

```{r, eval=F}
# If
if (3==2) {
  print("Condition 1 is TRUE!")
}

# If/Else
if (3==2) {
  print("Condition 2 is TRUE!")
} else {
  print("Condition 2 is FALSE!")
}
```


# Try it: for loops

```{r, eval=F}

my_vector <- c(2,4,6,8)
for (item in my_vector) {
  print(item)
}

```

# Try it: functions

```{r, eval=FALSE}
# Create a function that adds two numbers
add_numbers <- function(n1,n2) {
  print(n1 + n2)
}

# Invoke the function, supply the two arguments it needs
add_numbers(2,3)

# Redefine function with default values for arguments
add_numbers <- function(n1=5,n2=7) {
  print(n1 + n2)
}
add_numbers()
```

# Try it: create a data frame from scratch

- You can create a data frame from scratch using `data.frame()`

```{r, eval=F}
# People and birth country vectors, each of length 2
person <- c("Ding Liren", "Viswanathan Anand")
country <- c("China", "India")

# Combine into a data frame
X <- data.frame(Name = person, Birth_Country = country)
```

---

# Try it: inspect the data frame

```{r, eval=F}
## Print the dataset
X

# X is an object with class “data.frame” and type "list"
class(X)
typeof(X)

```

---

# Try it: subsetting a data frame

- Data frames are lists and can be subset using list subsetting operators

```{r, eval=F}
# Extract the first row of the data frame
X[1,]

# Extract the first column of the data frame
X[,1]
X[[1]]

# Data on people named "Viswanathan Anand"
X[X$Name == "Viswanathan Anand"]

# Extract the vector of birth countries
X$Birth_Country
```

---

# Read external data into a data frame

- Large data sets often come stored in external databases or plain text files

- `tidyverse` package includes functions to load most common formats into a data frame

```{r, eval = FALSE}
# Load package
library(readr)

# CSV "Comma Separated Values" Format
path <- "https://raw.githubusercontent.com/googletrends/data/master/20160819_OlympicSportsByCountries.csv"
trends <- read_csv(path, skip=2, n_max=100)

# Display the first few rows
head(trends)
```



