---
title: "Lab 10 - Cross-Validation"
author: Ives He
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## 0 load the package 
```{r}
library(boot)
```

## 1 create a data frame
```{r}
# Load the data
housing.df <- read.csv("/Users/iveshe/Library/Mobile Documents/com~apple~CloudDocs/Term 1/FIN 550/Lab/lab-10/BostonHousing.csv")

# First six rows 
head(housing.df)

# Column names 
names(housing.df)
```

## 2 leave-one-out cross-validation 

### 2.1 model 1: model MEDV as a cubic function of DIS 
```{r}
# fit a regression on the training set
glm1 <- glm(MEDV ~ poly(DIS, 3), data = housing.df)
summary(glm1)
# compute the leave-one-out cross-validation predictor error
loocv.err1 <- cv.glm(housing.df, glm1)
# cross-validated MSE 
loocv.err1$delta[1]
```

### 2.2 model 2: add a 4th degree polynomial in NOX as predictors, in addition to predictors in model 1
```{r}
# fit a regression on the training set 
glm2 <- glm(MEDV ~ poly(DIS, 3) + poly(NOX, 4), data = housing.df)
summary(glm2)
# compute the leave-one-out cross-validation predictor error
loocv.err2 <- cv.glm(housing.df, glm2)
# cross-validated MSE
loocv.err2$delta[1]
```

### 2.3 Model 3: add a 5th degree polynomial in CRIM as predictors, in addition to predictors in Model 2
```{r}
# fit a regression on the training set
glm3 <- glm(MEDV ~ poly(DIS, 3) + poly(NOX, 4) + poly(CRIM, 5), data = housing.df)
summary(glm3)
# compute the leave-one-out cross-validation predictor error
loocv.err3 <- cv.glm(housing.df, glm3)
# cross-validated MSE 
loocv.err3$delta[1]
```

### 2.4 identify the best a model according to the leave-one-out cross-validation method 
```{r}
# which model is best for predicting median home values? 
cat("LOOCV MSE Model 1:", loocv.err1$delta[1], "\n")
cat("LOOCV MSE Model 2:", loocv.err2$delta[1], "\n")
cat("LOOCV MSE Model 3:", loocv.err3$delta[1], "\n")

```

The best model is the one with the lowest MSE, so it is LOOCV MSE Model 2.

## 3 k-fold cross validation

### 3.1 model 1: model MEDV as a cubic function of DIS 
```{r}
# compute the 5-fold cross-validation predictor error
set.seed(1)
kfcv.err1 <- cv.glm(housing.df, glm1, K = 5)
# cross-validated MSE 
kfcv.err1$delta[1]
```

### 3.2 model 2: add a 4th degree polynomial in NOX as predictors, in addition to predictors in model 1 
```{r}
# compute the 5-fold cross-validation predictor error
kfcv.err2 <- cv.glm(housing.df, glm2, K = 5)
# cross-validated MSE 
kfcv.err2$delta[1]
```

### 3.3 Model 3: add a 5th degree polynomial in CRIM as predictors, in addition to predictors in Model 2
```{r}
# compute the 5-fold cross-validation predictor error
kfcv.err3 <- cv.glm(housing.df, glm3, K = 5)
# cross-validated MSE
kfcv.err3$delta[1]
```

### 3.4 identify the best a model according to the 5-fold cross-validation method 
```{r}
# which model is best for predicting median home values?  
cat("5-fold CV MSE Model 1:", kfcv.err1$delta[1], "\n")
cat("5-fold CV MSE Model 2:", kfcv.err2$delta[1], "\n")
cat("5-fold CV MSE Model 3:", kfcv.err3$delta[1], "\n")
```

The best model is the one with the lowest MSE, so it is 5-fold CV MSE Model 2.
