---
title: "Lab 04 - Data Exploration"
author: Ives He
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## 0 load the package 
```{r}
library(tidyverse)
```

## 1 create a tibble 

```{r}
# Load the data from firm.csv and manager.csv
firm_data <- read_csv("https://raw.githubusercontent.com/PYHkkk/FIN-550/main/Lab/Lab-04/lab-04-firm.csv")

# First six rows of firm data
head(firm_data)

# Column names for firm_data
colnames(firm_data)


```

## 2 calculate summary statistics 

```{r}
# 2.1 average net income   
average_net_income <- mean(firm_data$NI, na.rm = TRUE)
average_net_income


# 2.2 first decile of net income  
first_decile_net_income <- quantile(firm_data$NI, 0.1, na.rm = TRUE)
first_decile_net_income


# 2.3 summary of net income    
summary_net_income <- summary(firm_data$NI)
summary_net_income

```

## 3 compute counts and proportion of logical values    
```{r}
# 3.1 number of rows 
rows_condition <- sum(firm_data$DV > 0 & firm_data$NI > 10000, na.rm = TRUE)
rows_condition


# 3.2 proportion of rows where net income is positive
proportion_condition <- mean(firm_data$DV > 0 & firm_data$NI > 10000, na.rm = TRUE)
proportion_condition


```





## 4 summarize data with dpylr  

### 4.1 average and maximum net income 
```{r}
firm_data %>% 
  summarise(
    NI_mean = mean(NI, na.rm = TRUE),
    NI_max = max(NI, na.rm = TRUE),
    n_rows = n(),
    n_firms = n_distinct(TIC)
  )

```

### 4.2 average and minimum net income by firm 
```{r}
df_by_TIC <- firm_data %>% 
  group_by(TIC) %>% 
  summarise(
    NI_mean = mean(NI, na.rm = TRUE),
    NI_min = min(NI, na.rm = TRUE),
    n_rows = n()
  )
df_by_TIC

```

## 5 combine multiple operations with the pipe 

### 5.1 average net income by firm  
```{r}

firm_data %>% 
  group_by(TIC) %>% 
  summarise(NI_mean = mean(NI, na.rm = TRUE)) %>% 
  arrange(desc(NI_mean))

```

### 5.2 most recent return on assets by firm 
```{r}

firm_data %>% 
  mutate(ROA = NI / AT) %>% 
  group_by(TIC) %>% 
  arrange(desc(FYEAR)) %>% 
  summarise(recent_ROA = first(ROA))

```

## 6 combine firm fundamentals with executives’ compensation	

### 6.1 load executive compensation data 
```{r}
# Load the data  manager.csv
manager_data <- read_csv("https://raw.githubusercontent.com/PYHkkk/FIN-550/main/Lab/Lab-04/manager.csv")

# First six rows of manager data
head(manager_data)

# Column names for manager_data
colnames(manager_data)

```

### 6.2  frequency counts of firm
```{r}

table(firm_data$TIC)
table(manager_data$TICKER)

```
Answer: The firm_data dataset covers eight firms: AAPL, FB, FJTSY, GOOGL, IBM, INTC, MSFT, and ORCL, with each firm having four observations. The manager_data dataset covers seven firms: AAPL, FB, GOOGL, IBM, INTC, MSFT, and ORCL, but with different numbers of observations for each firm. The firm FJTSY is not present in the manager_data dataset. Therefore, the two datasets do not cover the exact same firms, as FJTSY appears in firm_data but not in manager_data. All the other firms are common between both datasets but with varying observation counts.

### 6.3 frequency counts of year
```{r}

table(firm_data$FYEAR)
table(manager_data$YEAR)

```
Answer: Both datasets cover the same fiscal years: 2015, 2016, 2017, and 2018. However, the number of observations per year differs between the two datasets: In firm_data, there are exactly 8 observations for each fiscal year. In manager_data, there are more observations, with 39 in 2015 and 2017, 40 in 2016, and 37 in 2018. Therefore, the two datasets cover the same fiscal years but with different numbers of observations for each year.

### 6.4 inner join 
```{r}
# Perform an inner join
merged <- inner_join(manager_data, firm_data, by = c("TICKER" = "TIC", "YEAR" = "FYEAR"))

# First six rows of merged data
head(merged)

# Column names
colnames(merged)

```

### 6.5 average salary and return on assets during 2017 and 2018 
```{r}

merged %>% 
  filter(YEAR %in% c(2017, 2018)) %>% 
  mutate(ROA = NI / AT) %>% 
  group_by(TICKER) %>% 
  summarise(
    SALARY_mean = mean(SALARY, na.rm = TRUE),
    ROA_mean = mean(ROA, na.rm = TRUE),
    n_rows = n()
  ) %>% 
  arrange(desc(ROA_mean))

```


