---
title: "Lecture 8"
---


# Try it: load libraries, turn off scientific notation
```{r, message=F}
library(tidyverse)

# scipen: a penalty for deciding whether to print fixed or exponential notation. 
# Positive values encourage fixed notation, negative encourage scientific notation
options(scipen=999)
```

---

# Try it: simulated dataset of housing values

```{r, message=F}
set.seed(2) # Set seed for replicability when generating random variables

# House ID (tibble, with 1000 obs)
housing_sim <- tibble(id = 1:1000)

# Number of rooms: use sample() to generate random integer between 2 and 7
housing_sim <- housing_sim %>% 
  mutate(rooms = sample(2:7, n(), replace = TRUE))

# Square feet: use runif() to generate random value between 200 and 400
housing_sim <- housing_sim %>% 
  mutate(sqft = rooms*runif(n(), 200, 400))

# error term: ranges from -5000 to 5000
housing_sim <- housing_sim %>% 
  mutate(error = 5000*runif(n(), -1, 1))
```



---

# Try it: simulated dataset of housing values


```{r b1, message=F, results=F}
# Home values = A*sqft + B*sqft^2 + C*rooms + error
housing_sim <- housing_sim %>% 
  mutate(value = 300*sqft - 0.05*sqft^2 + 10000*rooms + error)
head(housing_sim)
summary(housing_sim$value)
```


---

# Try it: univariate regression of values on sqft



```{r, message=F, results=F}

# Use lm() to regress housing values on sqft
# How does coefficient on sqft compare to the true value of 300?
```


---

# Try it: create new variable, sqft2, and add it to regression


- Is the coefficient on `sqft` closer to the correct value? Why is it not exact?


---

# Try it: use poly() function to add sqft2 to the regression


```{r b4, results = F}
# raw=T specifies raw polynomials
lm(value ~ poly(sqft, 2, raw = TRUE), data = housing_sim)
```

- Do you get the same answer as before?


---

# Try it: add rooms as a third predictor


---

# Try it: setting error term to zero

