---
title: "Lab 09 - Logistic Regression"
author: Ives He
output:
  html_document:
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## 0 load the package and suppress scientific notation 
```{r}
# load packages 
library(lattice)
library(ggplot2)
library(caret)
library(gains)
# first six rows

# not display scientific notation
options(scipen=999)
```

## 1 create a data frame
```{r}
# load the data
ebay.df <- read.csv("/Users/iveshe/Library/Mobile Documents/com~apple~CloudDocs/Term 1/FIN 550/Lab/lab-09/eBayAuctions.csv", stringsAsFactors = FALSE)
# variable names 
head(ebay.df)
names(ebay.df)
```

## 2 apply a function to each group 
```{r}
# mean of the binary outcome for each day of the week of auction close   
tapply(ebay.df$Competitive., ebay.df$endDay, mean)
# mean of the binary outcome for each auction category  
tapply(ebay.df$Competitive., ebay.df$Category, mean)
```

## 3 reduce levels in the day of week that the auction closed  
```{r}
# combine Sunday and Friday into a single category called Sun_Fri
ebay.df$endDay[ebay.df$endDay == "Sun"] <- "Sun_Fri"
ebay.df$endDay[ebay.df$endDay == "Fri"] <- "Sun_Fri"
```

## 4 reduce levels in auction categories 
```{r}
# combine Business/Industrial and Computers into a single category called Computers
ebay.df$Category[ebay.df$Category == "Business/Industrial"] <- "Computer"
# combine Antique/Art/Craft and Collectibles into a single category called collectibles
ebay.df$Category[ebay.df$Category == "Antique/Art/Craft"] <- "Collectibles"
```

## 5 convert Duration to a categorical or factor variable 
```{r}
# convert Duration from a numeric variable to a categorical or factor variable  
ebay.df$Duration <- as.factor(ebay.df$Duration)
# variable types 
str(ebay.df)
```

## 6 data partition
```{r}
# set seed for reproducing the partition 
set.seed(1)
# takes a sample of row numbers for the training set 
train.index <- sample(1:nrow(ebay.df), 0.6*nrow(ebay.df))
# training set 
train.df <- ebay.df[train.index, ]
# test set 
test.df <- ebay.df[-train.index, ]
```

## 7 fit a logistic regression model 
```{r,warning=FALSE}
# logistic regression using all the variables
reg <- glm(Competitive. ~ ., data = train.df, family = "binomial")
# summary table 
summary(reg)
```

## 8 generate predicted probabilities for records in the test set 
```{r}
# predicted probabilities 
pred <- predict(reg, test.df, type = "response")
# first six values 
head(pred)
```

## 9 confusion matrix 
```{r}
pred.class <- ifelse(pred > 0.5, 1, 0)
confusionMatrix(as.factor(pred.class), as.factor(test.df$Competitive.), positive = "1")
```

## 10 creating a gain table [optional]
```{r}
# gain table 
gain <- gains(test.df$Competitive., pred, groups = 10)
# cumulative percentage of competitive auctions 
gain$cume.pct.of.total
# cumulative number of auctions 
gain$cume.obs
```

## 11 plot a lift chart [optional]
```{r}
# plot the cumulative number of competitive auctions against the cumulative number of auctions
plot(c(0, gain$cume.obs), c(0, gain$cume.pct.of.total * sum(test.df$Competitive.)),
     xlab = "cumulative number of auctions", ylab = "cumulative number of competitive auctions", type = "l")
# add a baseline curve 
lines(c(0, dim(test.df)[1]), c(0, sum(test.df$Competitive.)), lty = 2)
```